---
output: html_document
editor_options: 
  chunk_output_type: console
---


### Network

#### EGA 
Traditionally, PCA and FA are the two most common factor analytic methods that were used to estimate the number of underlying factors in the observed data. 
EGA is a newly introduced method which applies the network modeling framework to estimate a network based on the relationships between the variables (estimate GLM with GLASSO) and subsequently identify the clusters of the variables in the network (community detection algorithms: louvain or walktrap). 

Simulation data has not only shown that EGA has higher accuracy than most PCA and FA methods

```{r, message=FALSE, warning=FALSE}
# Preprocessing
z <- effectsize::standardize(data[hrv_cols])
z <- z[rowSums(is.na(z)) < 0.2 * ncol(z), ]  # keep only participants with few nans
z[is.na(z)] <- 0 
```

##### EGA glasso
```{r, message=FALSE, warning=FALSE}
suppressMessages(library(EGAnet))

# Run EGA
# ega_glasso <- EGAnet::EGA(r, n = nrow(data), model = "glasso", algorithm = "louvain", plot.EGA = FALSE)
ega_glasso <- EGAnet::EGA(z, model = "glasso", algorithm = "louvain", plot.EGA = FALSE)


# Run Boot EGA
# ega_glasso_booth <- EGAnet::bootEGA(z, model = "glasso", algorithm = "louvain", plot.EGA = TRUE, iter = 500)
# ega_glasso_booth$summary.table
# # Frequency of each dimension solution
# ega_glasso_booth$frequency
# # Structural consistency
# stability <- EGAnet::dimensionStability(ega_glasso_booth)
# stability$dimension.stability$structural.consistency
# stability$item.stability$item.stability$all.dimensions
```

##### EGA TMFG
```{r, message=FALSE, warning=FALSE}
# Run EGA
# ega_TMFG <- EGAnet::EGA(r, n = nrow(data), model = "TMFG", algorithm = "louvain", plot.EGA = FALSE)
ega_TMFG <- EGAnet::EGA(z, model = "TMFG", algorithm = "louvain", plot.EGA = FALSE)

```

```{r, message=FALSE, warning=FALSE, cache = TRUE}
table_clusters$EGA_glasso <- ega_glasso$dim.variables[table_clusters$Index, "dimension"]
table_clusters$EGA_TMFG <- ega_TMFG$dim.variables[table_clusters$Index, "dimension"]
```


#### Centrality

```{r, message=FALSE, warning=FALSE, cache = TRUE}
graph_glasso <- EGAnet::ega_to_tidygraph(ega_glasso)  #from Dom's PR
graph_glasso <- tidygraph::tbl_graph(nodes = graph_glasso$nodes, edges = graph_glasso$edges, directed = FALSE) %>% 
  tidygraph::activate(nodes) %>%
  mutate(centrality = tidygraph::centrality_degree()) 

as.list(graph_glasso)$nodes %>% 
  group_by(dimension) %>% 
  arrange(desc(centrality)) %>% 
  slice_head(n=1) %>% 
  knitr::kable()

graph_TMFG <- EGAnet::ega_to_tidygraph(ega_TMFG)  #from Dom's PR
graph_TMFG <- tidygraph::tbl_graph(nodes = graph_TMFG$nodes, edges = graph_TMFG$edges, directed = FALSE) %>% 
  tidygraph::activate(nodes) %>%
  mutate(centrality = tidygraph::centrality_degree()) 

as.list(graph_TMFG)$nodes %>% 
  group_by(dimension) %>% 
  arrange(desc(centrality)) %>% 
  slice_head(n=1) %>% 
  knitr::kable()
```

#### Network Graph

```{r, message=FALSE, warning=FALSE, fig.width=figwidth, fig.height=figwidth}

graph_glasso %>% 
  ggraph::ggraph(layout = "graphopt") + 
    ggraph::geom_edge_arc(aes(colour = link, width = link), strength = 0.05) + 
    ggraph::geom_node_point(aes(colour = dimension, size = centrality)) + 
    ggraph::geom_node_text(aes(label = name), fontface = "bold", repel = FALSE) +
    ggraph::theme_graph() +
    ggraph::scale_edge_width(range = c(0.2, 3), guide = "none") +
    scale_color_material_d(palette = "rainbow", guide = "none") +
    ggraph::scale_edge_color_gradientn(limits = c(0, 1), colours=c("#607D8B", "#263238"), guide = "none") +
    scale_size_continuous(range = c(4, 20), guide = "none")

graph_TMFG %>% 
  ggraph::ggraph(layout = "graphopt") + 
    ggraph::geom_edge_arc(aes(colour = link, width = link), strength = 0.05) + 
    ggraph::geom_node_point(aes(colour = dimension, size = centrality)) + 
    ggraph::geom_node_text(aes(label = name), fontface = "bold", repel = FALSE) +
    ggraph::theme_graph() +
    ggraph::scale_edge_width(range = c(0.2, 3), guide = "none") +
    scale_color_material_d(palette = "rainbow", guide = "none") +
    ggraph::scale_edge_color_gradientn(limits = c(0, 1), colours=c("#607D8B", "#263238"), guide = "none") +
    scale_size_continuous(range = c(4, 20), guide = "none")
```
