---
title: '**Conceptual vs. Observed Structure of Heart Rate Variability (HRV) Indices**'
subtitle: "Data Analysis"
output:
  html_document:
    self_contained: false
    theme: cerulean
    highlight: pygments
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: no
    df_print: default
    code_folding: show
    code_download: yes
  word_document:
    reference_docx: utils/Template_Word.docx
    highlight: pygments
    toc: no
    toc_depth: 3
    df_print: default
    number_sections: yes
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '2'
    latex_engine: xelatex
editor_options:
  chunk_output_type: console
bibliography: utils/references.bib
csl: utils/apa.csl
---


<!-- 
!!!! IMPORTANT: run `source("utils/render.R")` to publish instead of clicking on 'Knit'
-->

```{r setup, warning=FALSE, message=TRUE, include=FALSE}
# Set up the environment (or use local alternative `source("utils/config.R")`)
source("https://raw.githubusercontent.com/RealityBending/TemplateResults/main/utils/config.R")  

fast <- FALSE  # Make this false to skip the chunks

# Set theme
ggplot2::theme_set(see::theme_modern())

source_rmd <- function(x){
  file <- knitr::purl(x, quiet=TRUE)
  source(file)
  if (file.exists(file)) file.remove(file)
}
```

# Introduction
```{r badges, echo=FALSE, message=TRUE, warning=FALSE, results='asis'}
# This chunk is a bit complex so don't worry about it: it's made to add badges to the HTML versions
# NOTE: You have to replace the links accordingly to have working "buttons" on the HTML versions
if (!knitr::is_latex_output() && knitr::is_html_output()) {
  cat("![Build](https://github.com/Tam-Pham/HRVStructure/workflows/Build/badge.svg)
      [![Website](https://img.shields.io/badge/repo-Readme-2196F3)](https://github.com/Tam-Pham/HRVStructure)
      [![Website](https://img.shields.io/badge/visit-website-E91E63)](https://realitybending.github.io/TemplateResults/)
      [![Website](https://img.shields.io/badge/download-.docx-FF5722)](https://github.com/RealityBending/TemplateResults/raw/main/word_and_pdf/SupplementaryMaterials.docx)
      [![Website](https://img.shields.io/badge/see-.pdf-FF9800)](https://github.com/RealityBending/TemplateResults/blob/main/word_and_pdf/SupplementaryMaterials.pdf)")
}
```

<!-- ABSTRACT -->

Heart Rate Variability (HRV) can be estimated using a myriad of mathematical indices, for which the lack of comparison renders results' interpretation and evaluation tedious. In this study, we assessed the relationship between 42 HRV metrics collected from Human recordings (n = 531) using various structure-analysis algorithms. We then applied a metaclustering approach that combines their results to obtain a robust and reliable view of the observed relationships. We found that HRV metrics can be clustered into 4 groups, representing .... From there, we derived recommendations on which indices to prioritize for parsimonious, yet comprehensive HRV-related data analysis and reporting.

<!-- introduce hrv -->

Heart Rate Variability (HRV), reflecting the heart's ability to effectively regulate and adapt to internal and external environmental changes, has been linked with many physical and mental health outcomes [e.g., cardiac complications, @laitio2007role, diabetes, @kudat2006heart, mood disorders, @bassett2016literature, cognitive functioning, @forte2019heart]. Conventionally, the various indices used in the assessment of HRV were broadly categorized based on the nature of their mathematical approach, in categories including *time-domain*, *frequency-domain*, and *nonlinear dynamics*.

<!-- is attributed to the dynamic interplay between the sympathetic and parasympathetic branches of the autonomic nervous system -->

<!-- main categories -->

<!-- DOM: summarize the 3 paragraphs below into 1; No need to present all the indices, just list some and say "for a comprehensive description of all indices, see Pham 2021. -->

Time-domain analysis presents the simplest and most straightforward method of quantifying HRV from the original normal (i.e., excluding abnormal beats such as ectopic beats) heartbeat intervals, henceforth referred to as NN intervals. Some commonly derived indices include *SDNN*, the standard deviation of all NN intervals, *SDSD*, the standard deviation of differences between adjacent NN intervals, *RMSSD*, the root mean square of the sum of successive differences of NN intervals, and *pNN50*, the percentage of adjacent NN intervals separated by more than 50ms.

While time-domain methods offers computational ease, they are unable to distinguish between the contributions of sympathetic and parasympathetic branches. Frequency-domain analysis, on the other hand, targets the assessment of these different regulatory mechanisms by investigating how the HRV power spectrum distributes across different frequency bands. The low frequency (*LF*; 0.04--0.15Hz) and high frequency (*HF*; 0.15--0.4 Hz) components, the two main indicators of ANS activity [@acharya2006heart], predominantly reflect sympathetic and parasympathetic activity respectively. The other two components measured at very low frequencies (*VLF*; 0.0033--0.04 Hz) and ultra low frequencies (*ULF* low frequency $\leq$ 0.003 Hz) are less clear in regards to their associated physiological mechanisms [@kleiger2005heart], though some evidence has linked the former to long-term regulatory mechanisms [@akselrod1981power] and the latter to circadian and metabolic activity [@shaffer2014healthy]. Other indices that fall under the frequency domain include derivatives of the aforementioned components, such as the ratio of *LF* to *HF* (*LF/HF*) power and their normalized (e.g., *LFn*, *HFn*) and natural logarithmic variants (e.g., *LnHF*).

Finally, drawn from concepts of non-linear dynamics and chaos theory [@golberger1996non], non-linear analysis was later introduced to better characterize the complex physiological mechanisms underlying HRV regulation. Prominent indices include measures obtained from a Poincaré plot where an ellipse is fitted to a scatterplot of each NN interval against its preceding one, specifically, the standard deviation of the short-term (*SD1*; width of ellipse) and long-term (*SD2* ; length of ellipse) NN interval variability [@brennan2001existing] as well as its corresponding ratio (*SD1*/*SD2*). Other non-linear indices that fall under this category, such as Detrended Fluctuation Analysis (*DFA*), multi-fractal *DFA* (*MF-DFA*) and correlation dimension (*CD*) account for the fractal properties of HRV, while entropy measures like approximate entropy (*ApEn*), sample entropy (*SampEn*), and multiscale entropy (*MSE*) quantify the amount of regularity in the HR time series [@voss2009methods]. 

<!-- overlap and duplicates in indices, aim of study -->

Despite the rising popularity of HRV analysis as a real-time, noninvasive technique for investigating health and disease, there are some shared similarities (and even overlaps) between the multitude of HRV indices that are not yet well understood. Early studies have investigated the relationships between time-domain and frequency-domain indices, showing that not only were *RMSSD* and *pNN50* strongly correlated with each other [above 0.9, @bigger1989comparison], they were also highly associated with *HF* power [@bigger1989comparison; @kleiger2005heart; @otzenberger1998dynamic], suggesting that these measures could be treated as surrogates for each other in assessing the parasympathetic modulation of HRV. This observation is warranted given that the former is computed from the differences across consecutive NN intervals, and hence, they reflect mainly high frequency oscillatory patterns in HR and are independent of long-term changes. On the other hand, *SDNN*, which has been thought to reflect both sympathetic and parasympathetic activity, is correlated to total power (*TP*) in HRV power spectrum [@bigger1989comparison]. Recent years also witnessed the emergence of debates regarding the traditional conceptualization of *SD1* and *SD2* as non-linear indices, particularly when @ciccone2017reminder proposed that *RMSSD* and *SD1* were mathematically equivalent. Many studies that report both of these short-term HRV indices independently often arrive at identical statistical results without addressing this equivalence [e.g., @rossi2015impact, @peng2015extraction, @leite2015correlation]. Additionally, other studies have also drawn similarities between *SD1/SD2* and *LF/HF* in their indexing of the balance between short- and long-term HRV [@brennan2002poincare; @guzik2007correlations]. 

Overall, there is a need for a greater data-driven understanding of the relationship between the multitude of HRV indices and their respective groupings. Although there have been attempts at investigating the association between HRV indices [e.g., @nguyen2019improving], there exists a large number of clustering analysis techniques and their usecases, limitations, and advantages are not clear. Moreover, these algorithms comprise of multiple parameters, for which there is no gold standard or clear guidelines for their specification. As such, choosing one method and presenting its solution as a definitive one can be misleading. Hence, the aim of this study is to explore the factor structure of HRV indices by using a more inclusive and probabilistic approach that we will refer to as meta-clustering.

<!--may be useful to have glossary of indices-->

# Methods

In total, the ECG data of 496 participants were extracted from 11 databases. The raw data as well as the entire analysis script (including details and additional analyses) can be found at this GitHub repository (<https://github.com/Tam-Pham/HRVStructure>).

Concept of **"METACLUSTERING"**, i.e., assessing the proximity based on multiple clustering solutions. Aka ENSEMBLE CLUSTERING.

## Databases

### Glasgow University Database

The GUDB Database [@howell2018high] contains ECGs from 25 subjects. Each subject was recorded performing 5 different tasks for two minutes (sitting, doing a maths test on a tablet, walking on a treadmill, running on a treadmill, using a hand bike). The sampling rate is 250Hz for all the conditions.

The script to download and format the database using the [**ECG-GUDB**](https://github.com/berndporr/ECG-GUDB) Python package by Bernd Porr can be found [**here**](https://github.com/neuropsychology/NeuroKit/blob/dev/data/gudb/download_gudb.py).

### MIT-BIH Arrhythmia Database

The MIT-BIH Arrhythmia Database [MIT-Arrhythmia; @moody2001impact] contains 48 excerpts of 30-min of two-channel ambulatory ECG recordings sampled at 360Hz and 25 additional recordings from the same participants including common but clinically significant arrhythmias (denoted as the `MIT-Arrhythmia-x` database).

The script to download and format the database using the can be found [**here**](https://github.com/neuropsychology/NeuroKit/blob/dev/data/mit_arrhythmia/download_mit_arrhythmia.py).


### MIT-BIH Normal Sinus Rhythm Database

This database includes 18 clean long-term ECG recordings of subjects. Due to memory limits, we only kept the second hour of recording of each participant.

The script to download and format the database using the can be found [**here**](https://github.com/neuropsychology/NeuroKit/blob/dev/data/mit_normal/download_mit_normal.py).


<!-- ### Lobachevsky University Electrocardiography Database -->

<!-- The Lobachevsky University Electrocardiography Database [LUDB; @kalyakulina2018lu] consists of 200 10-second 12-lead ECG signal records representing different morphologies of the ECG signal. The ECGs were collected from healthy volunteers and patients, which had various cardiovascular diseases. The boundaries of P, T waves and QRS complexes were manually annotated by cardiologists for all 200 records. -->

### Fantasia Database

The Fantasia database [@iyengar1996age] consists of twenty young and twenty elderly healthy subjects. All subjects remained in a resting state in sinus rhythm while watching the movie Fantasia (Disney, 1940) to help maintain wakefulness. The continuous ECG signals were digitized at 250 Hz. Each heartbeat was annotated using an automated arrhythmia detection algorithm, and each beat annotation was verified by visual inspection.


## Data Analysis

First, we correlated the HRV indices [using the `correlations` function from the *correlation* package, see @makowski2020methods] and removed indices that were perfected correlated with each other, before removing outliers based on the median absolute deviation from the median [see the `check_outliers` function in the *performance* R package; @ludecke2019performance]. Using a meta-clustering approach, multiple clustering methods were used to analyze the associations between the HRV indices, namely, factor analysis, k-means clustering, density-based spatial clustering of applications with noise (DBSCAN), hierarchical density-based spatial clustering of applications with noise (HBSCAN), and exploratory graph analysis (EGA). We then jointly considered the results obtained from these multiple methods, by estimating the probability for each pair of indices to be clustered together, in order to provide a more robust overview of the clustering results and facilitate a more thorough conceptual understanding of these patterns.

Data processing was carried out with R [@R-base] and the *easystats* ecosystem [@ludecke2019insight; @makowski2019bayestestr]. The raw data, as well as the full reproducible analysis script, including the solutions of each individual clustering method, are available in **Supplementary Materials 1**.



## Procedure

```{python, eval=FALSE, echo=FALSE}
import pandas as pd
import numpy as np
import neurokit2 as nk

# Load True R-peaks location
datafiles = [pd.read_csv("../../data/gudb/Rpeaks.csv"),
             pd.read_csv("../../data/mit_arrhythmia/Rpeaks.csv"),
             pd.read_csv("../../data/mit_normal/Rpeaks.csv"),
             pd.read_csv("../../data/fantasia/Rpeaks.csv")]

# Get results
all_results = pd.DataFrame()

for file in datafiles:
    for database in np.unique(file["Database"]):
        data = file[file["Database"] == database]
        for participant in np.unique(data["Participant"]):
            data_participant = data[data["Participant"] == participant]
            sampling_rate = np.unique(data_participant["Sampling_Rate"])[0]
            rpeaks = data_participant["Rpeaks"].values

            results = nk.hrv(rpeaks, sampling_rate=sampling_rate)
            results["Participant"] = participant
            results["Database"] = database
            results["Recording_Length"] = rpeaks[-1] / sampling_rate / 60

            all_results = pd.concat([all_results, results], axis=0)

all_results.to_csv("data/data.csv", index=False)
```

# Results

```{r, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(easystats)

data <- read.csv("data/data_combined.csv", stringsAsFactors = FALSE) %>% 
  select(-HRV_ULF, -HRV_VLF) %>%  # Empty
  # filter(Database != "LUDB") %>% # too short recordings, many indices didn't converge
  setNames(stringr::str_remove(names(.), "HRV_")) %>% 
  mutate_all(function(x) {
    x[is.infinite(x)] <- NA
    return(x) })
```


```{r warning=FALSE, message=TRUE, results='asis'}
cat("This study includes", 
    nrow(data), 
    "participants from", 
    length(unique(data$Database)), 
    "databases.")
```


```{r convenience_chunk_for_local_running, include = FALSE, eval = FALSE}
library(patchwork)
library(tidyverse)
library(easystats)


# # To run things locally: CTRL + ALT + SHIFT + P, and then the next two lines
source_rmd("0_ConvenienceFunctions.Rmd")
source_rmd("1_Preprocessing.Rmd")
source_rmd("3_Dimensions.Rmd")
source_rmd("4_Clustering.Rmd")

find_cache <- function(pattern = "") {
  files <- paste0("index_cache/html/", list.files("index_cache/html/", pattern = "\\.rdx$"))
  files <- files[stringr::str_detect(files, pattern)]
  stringr::str_remove(files, ".rdx")
}

lapply(find_cache(), lazyLoad, environment())
```



## Preprocessing

```{r child='0_ConvenienceFunctions.Rmd'}
```


```{r child='1_Preprocessing.Rmd'}
```

## Effect of Recording Length

```{r child='2_RecordingLength.Rmd'}
```


## Dimensional Structure

```{r child='3_Dimensions.Rmd'}
```

## Cluster Structure

```{r child='4_Clustering.Rmd'}
```

## Network Structure

```{r child='5_Network.Rmd'}
```

## Summary

```{r child='6_Summary.Rmd'}
```



# Discussion

<!-- which indices most central -->

# References
